Incremental λ-Calculus in Cache-Transfer Style

### 2017-07-25 popl18/soundness.tex
Spacing/layout/wordsmithing for formalization chapter
Page breaks were ugly, so I ended up revising paragraphs that ended in short
lines to save them (and try to improve text).

In this section, we outline definitions and main lemmas needed to prove CTS
conversion sound.
The proof is based on a
mostly straightforward simulation in lock-step, but two subtle points emerge.
First, we must relate $\source$ environments that do not contain caches, with
$\target$ environments that do.
Second, while evaluating CTS derivatives, the evaluation environment mixes
caches from the base computation and updated caches computed by the derivatives.

\paragraph{Evaluation commutes with CTS conversion}

@@

Figure~\ref{fig:differentiation-and-static-caching-continued}
extends CTS translation to values, change values, environments and
change environments. CTS translation of base terms commutes with our semantics, as shown by
next lemma:

\begin{lemma}[Evaluation commutes with CTS translation]
  For all $\sectx, \sterm$ and $\sclosedvalue$, such that
  $\seval\sectx\sterm{}\sclosedvalue$, and
  for all $\tcache$, there exists $\tcachedclosedvalues$,
  $\seval
  {\compile{\sectx}}{\compileterm\tcache\sterm}
  {}
  {(\compile{\sclosedvalue}, \tcachedclosedvalues)}$.
\end{lemma}

@@

%\pg{This consumes and produces change environments!}
As a fix, we enrich~$\compile\iectx$ with the values of a cache~$\tcache$, using
the pair of judgments $\envwithcache\tdectx\tcache{\tdectx'}$ (for $i = 1, 2$) defined in \cref{fig:envwithcache}.
Judgment~$\envwithcache\tdectx\tcache{\tdectx'}$ (for $i = 1, 2$) is read ``Target
change environment~$\tdectx'$ extends~$\tdectx$ with the original (for $i = 1$)
(or updated, for $i = 2$) values of
cache~$\tcache$.''
Since $\tcache$ is essentially a list of variables containing no values, cache
values in $\tdectx'$ must be computed from $\env\tdectx$.

\begin{lemma}[Evaluation commutes with CTS differentiation]
  Let $\tcache$ be such that
  $(\tcache, \_) = \compileterm{\bullet}{\sterm}$.
  For all $\iectx, \sterm$ and $\icloseddvalue$,
  if
  $\seval{\iectx}{\iderive\sterm}{}{\icloseddvalue}$,
  and $\oldenvwithcache{\compile\iectx}{\tcache}{\tdectx}$,
  then
  $\seval{\tdectx}{\derive{\tupdcache}{\sterm}}{}{(\compile\icloseddvalue, \tcachedclosedvalues)}$.
\end{lemma}

The proof of this lemma is not immediate, since during the
evaluation of~$\derive{\tupdcache}{\sterm}$ the new caches replace the
old caches. In our Coq development, we enforce a physical separation
between the part of the environment containing old caches and the
one containing new caches, and we maintain the invariant that the
second part of the environment corresponds to the remaining part of the
term.

>>>>

In this section, we outline definitions and main lemmas needed to prove CTS
conversion sound.
The proof is based on a
mostly straightforward simulation in lock-step, but two subtle points emerge.
First, we must relate $\source$ environments that do not contain caches, with
$\target$ environments that do.
Second, while evaluating CTS derivatives, the evaluation environment mixes
caches from the base computation and updated caches computed by the derivatives.
\paragraph{Evaluation commutes with CTS conversion}

@@

Figure~\ref{fig:differentiation-and-static-caching-continued}
extends CTS translation to values, change values, environments and
change environments. CTS translation of base terms commutes with our semantics:
\begin{lemma}[Evaluation commutes with CTS translation]
  For all $\sectx, \sterm$ and $\sclosedvalue$, such that
  $\seval\sectx\sterm{}\sclosedvalue$, and
  for all $\tcache$, there exists $\tcachedclosedvalues$,
  $\seval
  {\compile{\sectx}}{\compileterm\tcache\sterm}
  {}
  {(\compile{\sclosedvalue}, \tcachedclosedvalues)}$.
\end{lemma}

@@

%\pg{This consumes and produces change environments!}
As a fix, we enrich~$\compile\iectx$ with the values of a cache~$\tcache$, using
the pair of judgments $\envwithcache\tdectx\tcache{\tdectx'}$ (for $i = 1, 2$) defined in \cref{fig:envwithcache}.
Judgment~$\envwithcache\tdectx\tcache{\tdectx'}$ (for $i = 1, 2$) is read ``Target
change environment~$\tdectx'$ extends~$\tdectx$ with the original (for $i = 1$)
(or updated, for $i = 2$) values of
cache~$\tcache$.''
Since $\tcache$ is essentially a list of variables containing no values, cache
values in $\tdectx'$ must be computed from $\env\tdectx$.
\begin{lemma}[Evaluation commutes with CTS differentiation]
  Let $\tcache$ be such that
  $(\tcache, \_) = \compileterm{\bullet}{\sterm}$.
  For all $\iectx, \sterm$ and $\icloseddvalue$,
  if
  $\seval{\iectx}{\iderive\sterm}{}{\icloseddvalue}$,
  and $\oldenvwithcache{\compile\iectx}{\tcache}{\tdectx}$,
  then
  $\seval{\tdectx}{\derive{\tupdcache}{\sterm}}{}{(\compile\icloseddvalue, \tcachedclosedvalues)}$.
\end{lemma}
The proof of this lemma is not immediate, since during the
evaluation of~$\derive{\tupdcache}{\sterm}$ the new caches replace the
old caches. In our Coq development, we enforce a physical separation
between the part of the environment containing old caches and the
one containing new caches, and we maintain the invariant that the
second part of the environment corresponds to the remaining part of the
term.

### 2017-07-25 popl18/source-language.tex
Spacing/layout/wordsmithing for formalization chapter
Page breaks were ugly, so I ended up revising paragraphs that ended in short
lines to save them (and try to improve text).

Our syntax for base terms~$\sterm$ represents $\lambda$-lifted
programs in A'-normal form (\cref{sec:motivating-example}).
We write $\many\tx$ for a sequence of variables of some unspecified length
$\tx_1, \tx_2, \ldots, \tx_m$.
A term can be either a bound variable~$\tx$, or a {\bf let}
binding of $\ty$ in subterm~$\sterm$ to either a new
tuple~$\stuple{\many\tx}$ ($\slet{\ty = \stuple{\many\tx}}\sterm$), or
the result of calling function~$\tf$ on argument~$\tx$
($\slet{\ty = \sapp{\tf}{\tx}}{\sterm}$).  Both~$\tf$ and~$\tx$ are
variables to be looked up in the environment.  Terms cannot contain
$\lambda$-abstractions as they have been $\lambda$-lifted to top-level
definitions, which we encode as closures in the initial environments.
%
Unlike standard ANF we add no special syntax for function
calls in tail position (see \cref{sec:tail-calls} for a discussion
about this limitation).
%
We often need to inspect the result of
a function call~$\tf\,\tx$, which is not a valid term in our syntax.
To alleviate this, we write~``$\scall\tf\tx$''
for ``$\slet{\ty = \tf\;\tx}\ty$'' where $\ty$ is
chosen fresh.

>>>>

Our syntax for base terms~$\sterm$ represents $\lambda$-lifted
programs in A'-normal form (\cref{sec:motivating-example}).
We write $\many\tx$ for a sequence of variables of some unspecified length
$\tx_1, \tx_2, \ldots, \tx_m$.
A term can be either a bound variable~$\tx$, or a {\bf let}
binding of $\ty$ in subterm~$\sterm$ to either a new
tuple~$\stuple{\many\tx}$ ($\slet{\ty = \stuple{\many\tx}}\sterm$), or
the result of calling function~$\tf$ on argument~$\tx$
($\slet{\ty = \sapp{\tf}{\tx}}{\sterm}$).  Both~$\tf$ and~$\tx$ are
variables to be looked up in the environment.  Terms cannot contain
$\lambda$-abstractions as they have been $\lambda$-lifted to top-level
definitions, which we encode as closures in the initial environments.
%
Unlike standard ANF we add no special syntax for function
calls in tail position (see \cref{sec:tail-calls} for a discussion
about this limitation).
%
We often inspect the result of
a function call~$\tf\,\tx$, which is not a valid term in our syntax.
To enable this, we write~``$\scall\tf\tx$''
for ``$\slet{\ty = \tf\;\tx}\ty$'' where $\ty$ is
chosen fresh.

### 2017-07-25 popl18/target-language.tex
Spacing/layout/wordsmithing for formalization chapter
Page breaks were ugly, so I ended up revising paragraphs that ended in short
lines to save them (and try to improve text).

In this section, we present the target language of a transformation
that extends static differentiation with CTS conversion.
As said earlier, the functions of $\target$ not only compute their
output but they also returns a cache containing the intermediate
values that have contributed to that output. This cache is transmitted
to derivatives to build changes on top of intermediate values with no
recomputation and derivatives update the caches according to their
input changes.

\paragraph{Syntax}
The syntax of $\target$ is defined in
Figure~\ref{fig:target-definition-syntax}. The base terms of $\target$
follow the same shape as the $A$-normal $\lambda$-lifted terms of
$\source$ except that a $\bf let$-binding for a function
application~$\tf\,\tx$ now introduces an extra \textit{cache
  identifier}~$\tcacheid\ty\tf\tx$ in replacement for the
output~$\ty$. The syntax of cache identifiers is non-standard: it can
be seen as a triple that refers to the value identifiers $\tf$, $\tx$
and $\ty$. Hence, an $\alpha$-renaming of one of these three
identifiers must refresh the cache identifier accordingly. The syntax
$(\tx, \tcache)$ for the result term makes explicit that a cache
$\tcache$ is systematically returned.
% A similar alpha-renaming issue must be made about the formal
% cache argument of caching-derivatives. In the Coq development,
% the positional approach to binders makes the problem disappear.

@@

The syntax for cache updates is almost the same as the one for caches
except each value identifier~$\tx$ of the input cache is adjusted with
its corresponding change~$\tdx$.

@@

Rule~\rulename{TDResult} returns the final change value of a
computation as well as a updated cache resulting from the evaluation
of the cache update term~$\tupdcache$.
%
The rule~\rulename{TDTuple} is similar to its counterpart in the
source language, except that no tuple is built for~$\ty$ as it has
already been pushed in the environment by the cache.

@@

If $\tdf$ is bound to a nil change and $\tf$ is bound to a primitive, the
rule~\rulename{TDPrimitiveNil} applies. The derivative of the
primitive is invoked with the value of $\tx$, its change value and the
cache of the original call to this primitive. The semantics of this
derivative is given by a builtin function~$\tddelta\sprim\param$, as
in the source language.

>>>>

In this section, we present the target language of a transformation
that extends static differentiation with CTS conversion.
As said earlier, the functions of $\target$ compute both their
output and a cache, which contains the intermediate
values that contributed to that output. Derivatives receive this cache and use
it to compute changes without recomputing intermediate values; derivatives also
update the caches according to their input changes.

\paragraph{Syntax}
The syntax of $\target$ is defined in
Figure~\ref{fig:target-definition-syntax}. Base terms of $\target$
follow again $\lambda$-lifted A'NF, like
$\source$, except that a $\bf let$-binding for a function
application~$\tf\,\tx$ now binds an extra \textit{cache
  identifier}~$\tcacheid\ty\tf\tx$ instead of
output~$\ty$. Cache identifiers have non-standard syntax: it can
be seen as a triple that refers to the value identifiers $\tf$, $\tx$
and $\ty$. Hence, an $\alpha$-renaming of one of these three
identifiers must refresh the cache identifier accordingly.
Result terms explicitly return cache $\tcache$ through
syntax $(\tx, \tcache)$.
% A similar alpha-renaming issue must be made about the formal
% cache argument of caching-derivatives. In the Coq development,
% the positional approach to binders makes the problem disappear.

@@

The syntax for cache updates resembles the one for caches,
but each value identifier~$\tx$ of the input cache is updated with
its corresponding change~$\tdx$.

@@

Rule~\rulename{TDResult} returns the final change value of a
computation as well as a updated cache resulting from the evaluation
of the cache update term~$\tupdcache$.
%
Rule~\rulename{TDTuple} resembles its counterpart in the
source language, but the tuple for~$\ty$ is not built as it has already been
pushed in the environment by the cache.

@@

If $\tdf$ is bound to a nil change and $\tf$ is bound to primitive $p$, the
rule~\rulename{TDPrimitiveNil} applies. The derivative of $\tprim$
is invoked with the value of $\tx$, its change value and the
cache of the original call to $\tprim$. The semantics of $\tprim$'s
derivative is given by builtin function~$\tddelta\sprim\param$, as
in the source language.
